---
title: 'Statistical Inference Course Project Part 1'
author: 'Carlos S. Pérez'
output:
  pdf_document: default
---

# Simulation Exercises
This project investigate the exponential distribution in R and compare it with the Central Limit Theorem.

The exponential distribution can be simulated in R with `rexp(n, lambda)` where
`lambda` $\lambda$ is the rate parameter. The mean of exponential distribution is 
$1/\lambda$ and the standard deviation is also $1/\lambda$. For this simulation,
we set $\lambda=0.2$ for all the simulations. 

The following code generates a thousand simulated averages of 40 exponentials.

```{r}
library(ggplot2)
#Make reproducible
set.seed(881014)

lambda <- 0.2

num_simulations <- 1000

sample_size <- 40

#all simulations
simulations <- matrix(rexp(num_simulations*sample_size,rate=lambda),
               num_simulations, sample_size)

#Average
exps_means <- rowMeans(simulations)
```

### Sample Mean versus Theoretical Mean

The next plot shows the distribution of the sample average (i.e. the simulated mean of 40 exponentials), since the sample mean is distributed in a normal fashion we expect that its mean (the mean of the sample mean distribution) will be close $1/\lambda$.

```{r,  fig.width=5, fig.height=3}
# plot the histogram of sample mean with legend
hist(exps_means, breaks=50, prob=TRUE,
     main="Sample Mean Distribution",xlab="")
lines(density(exps_means))
abline(v=1/lambda, col="blue")
xfit <- seq(min(exps_means), max(exps_means), length=100)
yfit <- dnorm(xfit, mean=1/lambda, sd=(1/lambda/sqrt(sample_size)))
lines(xfit, yfit, pch=22, col="blue", lty=2)
legend('topright', c("Simulated", "Theoretical"), lty=c(1,2), col=c("black", "blue"))
```

The plot shows that the distribution of sample means is centered at `r mean(exps_means)`
and the theoretical center of the distribution is $\lambda^{-1}$ = `r 1/lambda` indicated with blue solid line.

### Sample Variance versus Theoretical Variance

If a variance comparison is made we have that the **sample means variance** is `r var(exps_means)` while the **theoretical variance**
of the distribution is $\sigma^2 / n = 1/(\lambda^2 n) = 1/(0.04 \times 40)$ =
`r 1/(0.04 * 40)`. 

### Simulated Distribution 

The next figure compare the distribution of averages of 40 exponentials, suggesting that it is very close to a normal distribution.

```{r, fig.width=5, fig.height=3}
#plot of distribution difference
qqnorm(exps_means); qqline(exps_means)
```

