---
title: "Transmission Effect on Miles Per Galon (MPG)"
output:
  pdf_document: default
---

# Motor Trend, a magazine about the automobile industry. 

### Problem

Looking at a data set of a collection of cars, interested in exploring the relationship between a set of variables and miles per gallon (MPG) (outcome variable). 

They are particularly interested in the following two questions

- “Is an automatic or manual transmission better for MPG”

- "Quantify the MPG difference between automatic and manual transmissions"

### Data

First mtcars data set is loaded at start from R base

### Analysis 

#### Exploration 


First we clean and preparate the data
```{r, fig.cap = "Figure 1", echo = TRUE}
library(ggplot2)
library(corrplot)

mtcars2<-mtcars

#factor labeling
mtcars$gear <- factor(mtcars$gear,levels=c(3,4,5),
  	labels=c("3gears","4gears","5gears")) 
mtcars$am <- factor(mtcars$am,levels=c(0,1),
  	labels=c("Automatic","Manual")) 
mtcars$cyl <- factor(mtcars$cyl,levels=c(4,6,8),
   labels=c("4cyl","6cyl","8cyl")) 

summary(mtcars)
```

Figure number 2 shows the distribution of miles per galon in relation with horsepower, cylinders, gears and transmission. 
The plot shows a concentration over the upper left corner for 4 cyl where manual is the major part focusing around 27, while in 8 cyl and 3gears the automatic type is clustered around 15 miles. 

```{r, fig.cap = "Figure 2", echo = TRUE}
#mpg vs hp with transmission, facet cyl vs gear

qplot(hp, mpg, data=mtcars, shape=am, color=am, 
   facets=gear~cyl, size=I(3),
   xlab="Horsepower", ylab="Miles per Gallon") 
```


The next plot shows a basic exploratory analysis in terms of correlation matrix, where can be seen that mpg is strongly related to cyl, disp, hp and wt.
```{r, fig.cap = "Figure 5", echo = TRUE}

cor_plot<-corrplot.mixed(cor(mtcars2))

```


#### Regression 
In the next plot it is done a piecewise  regression by type and some relations are contrasted.
```{r, fig.cap = "Figure 3", echo = TRUE, warning=FALSE}

# Separate regressions of mpg on weight for each number of cylinders and transmission
qplot(wt, mpg, data=mtcars, geom=c("point", "smooth"), 
   method="lm", formula=y~x, color=cyl,shape=am, 
   main="Regression of MPG on Weight", 
   xlab="Weight", ylab="Miles per Gallon")

```

Next 3 models are done: one only with am variables, another with all variables and a lastone with step wise selection for predictors.
```{r, echo = FALSE}

# univariate model
am_model <- lm(mpg ~ am, data = mtcars)

# all vars model
all_model = lm(mpg ~ ., data = mtcars)

# step wise regression
step_model <- step(lm(mpg ~ ., data = mtcars), trace = 0)

# coefficients
summary(am_model)
summary(all_model)
summary(step_model)

#anova results
anova(am_model, step_model,all_model)

```
As can be observed the anova results indicate that the step wise model has a better performance. 
The figure number 4 show the basic plots for a regression fit. The one on residuals and q-q plot show that residuals have a very-close-to normal distribution.   

```{r, fig.cap = "Figure 4", echo = TRUE}

par(mfrow=c(2,2))
# Reg plots
plot(step_model)

```

For testing the global effect of transmission type over mpg hypothesis testing is done.

```{r, echo = FALSE}

#F test for variances ratio
auto<-mtcars[mtcars$am=='Automatic','mpg']
man<-mtcars[mtcars$am=='Manual','mpg']

var.test(auto,man)

#t test for difference of means 

t.test(mpg ~ am, data = mtcars)

```
